---
title: "MASTERING SHINY - Notes"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# MASTERING SHINY 
## Hadley Wickham
### 2020
[Book](https://mastering-shiny.org/index.html)


# 1. Packages: List of needed pacakges:

"gapminder", "ggforce", "gh", "globals", "openintro", "profvis", 
"RSQLite", "shiny", "shinycssloaders", "shinyFeedback", "shinythemes", "testthat", "thematic", "tidyverse", "vroom", "waiter", "xml2","zeallot" 
 - install.packages(c(list of package))
 
 - library(package)
 
 - library(pacman)
 
 - p_load(
  gapminder, ggforce, gh, globals, openintro, profvis, RSQLite, shiny,
  shinycssloaders, shinyFeedback, shinythemes, testthat, thematic,
  tidyverse, vroom, waiter, xml2, zeallot)
  
  
# 2. Shiny app:

**There 2 ways to create a shiny app:**

  - new dir + app.R file in 1 step: File > New Project > dir > Shiny Web Application 
  - file > New file > app.r (app.R file ready, use boilerplate:  “shinyapp” +  Shift+Tab)y typing “shinyapp” and pressing Shift+Tab.



**There are a few ways you can run this app:**

  - Click the Run App (Figure 1.1) button in the document toolbar.
  - Use a keyboard shortcut: Cmd/Ctrl + Shift + Enter.
  - If Not in RStudio: 
      - (source()) the whole document, or 
      - call shiny::runApp() with the path to the dir containing app.R


**To stop the app, choose one of these options:**

  - Click the stop sign icon on the R console toolbar.
  - Click on the console, then press Esc (Ctrl + C if using RStudio).
  - Close the Shiny app window.
  
  
# 3. Basic UI:

**Shiny extension package:**

A comprehensive, actively-maintained list of other packages  maintained by Nan Xiao can be found [here](https://github.com/nanxstats/awesome-shiny-extensions).


### 3.1 Inputs

- Common structure (for all :

  - inputID: abcd_1234 (no space-dash-period-special char,  must be unique)
  
  - label: human-readable label
  
  - value: set default value
  
- When creating an input, I recommend supplying the *inputId* and *label* arguments by position, and all other arguments by name:

```{r}
sliderInput("age_id", "Patient age", value = 50, min = 0, max = 100)
```


- Examples: 

  - Free text: *textInput()* - *passwordInput()* - *textAreaInput()*
  - Numeric inputs: *numericInput()* - *sliderInput()*. [Customizing](https://shiny.rstudio.com/articles/sliders.html)
  - Dates: *dateInput()* - *dateRangeInput()*. 
  - Limited choices: *selectInput()* - *radioButtons()* - *checkboxInput()* - *checkboxGroupInput()*.
  - File uploads: *fileInput()*
  - Action buttons: *actionButton()* - *actionLink()*


### 3.2 Outputs

**Outputs** in the UI create placeholders, later filled by the server function. Outputs take a unique ID as their first arg (outputId): 

ID "plot" => output$plot.

Each output func on the front end is coupled with a render func in the back end. There are three main types of output: text, tables, and plots. 

 - Text: 
    - *textOutput()* => *renderText()* 
    - *verbatimTextOutput()*  => *renderPrint()*
    - Handy package: [reactable](https://glin.github.io/reactable/index.html) to costumize tables.
 
 - Tables: 2 options for displaying df tables:
    - static (showing all): *tableOutput()* => *renderTable()* 
    - dynamic (fixed # of rows w/ controls): *dataTableOutput()* => *renderDataTable()*
  
- Plots: You can display any type of R graphic:
    - base, ggplot2: *plotOutput()* => *renderPlot()*
    - Plotly: *plotlyOutput()* => *renderPlotly()*
    
  
  
# 4. Basic reactivity

## 4.1 The server function (input, output, session)

 - Input:
    - The `input` argument is a list-like object that contains all the input data sent from the browser, named according to the input ID.
    - `input` objects are read-only. If you attempt to modify an input inside the server function, you’ll get an error.
    - important: `input` is selective about who is allowed to read it. To read from an input, you must be in a reactive context created by a function like `renderText()` or `reactive()`.

 - Output:
    - `output` is also a list-like object named according to the output ID. 
    - The main difference is that you use it for sending output instead of receiving input. 
    - You always use the output object in concert with a render function, as in:
    



```{r}
ui <- fluidPage(
  textOutput("greeting")
)

server <- function(input, output, session){
  output$greeting -> renderText('Hello')
}

```

Important:

 - The ID is quoted in the UI, but not in the server.
 - The render function does two things:
    - sets up a special reactive context that automatically tracks what inputs the output uses.
    - converts the output of your R code into HTML suitable for display on a web page.
 - Like the input, the output is picky about how you use it. You’ll get an error if:
    - You forget the render function: `output$greeting <- "Hello"`
    - You attempt to read from an output: ` message("The greeting is ", output$greeting)`
    
## 4.2 Reactive programming

 - Imperative vs declarative programming
    - Imperative progr: specific command, carried out immediately (loading data, wranfling, plotting, saving...)
    - Decrative progr: higher-level goals or important constraints (shiny)
 - Laziness: app wont run if mispelling (ne error flag)! Double check that your UI and server functions are using the same identifiers.
 - The reactive graph: *gtg* has a reactive dependency on *name*. see reactlog (ch14)
 - Reactive expressions: tool that reduces duplication in your reactive code by introducing additional nodes into the reactive graph `rct_expr()`.
 - Execution order:
    - The order in which reactive code is run is determined only by the reactive graph, not by its layout in the server function.
    - Make sure that reactive expressions and outputs only refer to things defined above, not below11. This will make your code easier to understand.
 
## 4.3 Reactive expressions

  - **producers** will refer to reactive inputs and expressions, 
  - **consumers** will refer to reactive expressions and outputs

### 4.3.1 Motivation



## 4.4 Controlling timing of evaluation

## 4.5 Observers
