---
title: "MASTERING SHINY - BASICS"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# MASTERING SHINY 
## Hadley Wickham
### 2020
[Book](https://mastering-shiny.org/index.html)


# 1. Packages: List of needed pacakges:

"gapminder", "ggforce", "gh", "globals", "openintro", "profvis", 
"RSQLite", "shiny", "shinycssloaders", "shinyFeedback", "shinythemes", "testthat", "thematic", "tidyverse", "vroom", "waiter", "xml2","zeallot" 
 - install.packages(c(list of package))
 
 - library(package)
 
 - library(pacman)
 
 - p_load(
  gapminder, ggforce, gh, globals, openintro, profvis, RSQLite, shiny,
  shinycssloaders, shinyFeedback, shinythemes, testthat, thematic,
  tidyverse, vroom, waiter, xml2, zeallot)
  
  
# 2. Shiny app:

**There 2 ways to create a shiny app:**

  - new dir + app.R file in 1 step: File > New Project > dir > Shiny Web Application 
  - file > New file > app.r (app.R file ready, use boilerplate:  “shinyapp” +  Shift+Tab)y typing “shinyapp” and pressing Shift+Tab.



**There are a few ways you can run this app:**

  - Click the Run App (Figure 1.1) button in the document toolbar.
  - Use a keyboard shortcut: Cmd/Ctrl + Shift + Enter.
  - If Not in RStudio: 
      - (source()) the whole document, or 
      - call shiny::runApp() with the path to the dir containing app.R


**To stop the app, choose one of these options:**

  - Click the stop sign icon on the R console toolbar.
  - Click on the console, then press Esc (Ctrl + C if using RStudio).
  - Close the Shiny app window.
  
  
# 3. Basic UI:

**Shiny extension package:**

A comprehensive, actively-maintained list of other packages  maintained by Nan Xiao can be found [here](https://github.com/nanxstats/awesome-shiny-extensions).


### 3.1 Inputs

- Common structure (for all :

  - inputID: abcd_1234 (no space-dash-period-special char,  must be unique)
  
  - label: human-readable label
  
  - value: set default value
  
- When creating an input, I recommend supplying the *inputId* and *label* arguments by position, and all other arguments by name:

```{r}
sliderInput("age_id", "Patient age", value = 50, min = 0, max = 100)
```


- Examples: 

  - Free text: *textInput()* - *passwordInput()* - *textAreaInput()*
  - Numeric inputs: *numericInput()* - *sliderInput()*. [Customizing](https://shiny.rstudio.com/articles/sliders.html)
  - Dates: *dateInput()* - *dateRangeInput()*. 
  - Limited choices: *selectInput()* - *radioButtons()* - *checkboxInput()* - *checkboxGroupInput()*.
  - File uploads: *fileInput()*
  - Action buttons: *actionButton()* - *actionLink()*


### 3.2 Outputs

**Outputs** in the UI create placeholders, later filled by the server function. Outputs take a unique ID as their first arg (outputId): 

ID "plot" => output$plot.

Each output func on the front end is coupled with a render func in the back end. There are three main types of output: text, tables, and plots. 

 - Text: 
    - *textOutput()* => *renderText()* 
    - *verbatimTextOutput()*  => *renderPrint()*
    - Handy package: [reactable](https://glin.github.io/reactable/index.html) to costumize tables.
 
 - Tables: options for displaying df tables:
    - static (showing all): *tableOutput()* => *renderTable()* 
    - dynamic (fixed # of rows w/ controls): *dataTableOutput()* => *renderDataTable()*
    - reactable: (costumizable): `reactableOutput()` => `renderReactable()`
  
- Plots: You can display any type of R graphic:
    - base, ggplot2: *plotOutput()* => *renderPlot()*
    - Plotly: *plotlyOutput()* => *renderPlotly()*
    
  
  
# 4. Basic reactivity

## 4.1 The server function (input, output, session)

 - Input:
    - The `input` argument is a list-like object that contains all the input data sent from the browser, named according to the input ID.
    - `input` objects are read-only. If you attempt to modify an input inside the server function, you’ll get an error.
    - important: `input` is selective about who is allowed to read it. To read from an input, you must be in a reactive context created by a function like `renderText()` or `reactive()`.

 - Output:
    - `output` is also a list-like object named according to the output ID. 
    - The main difference is that you use it for sending output instead of receiving input. 
    - You always use the output object in concert with a render function, as in:
    



```{r}
ui <- fluidPage(
  textOutput("greeting")
)

server <- function(input, output, session){
  output$greeting -> renderText('Hello')
}

```

Important:

 - The ID is quoted in the UI, but not in the server.
 - The render function does two things:
    - sets up a special reactive context that automatically tracks what inputs the output uses.
    - converts the output of your R code into HTML suitable for display on a web page.
 - Like the input, the output is picky about how you use it. You’ll get an error if:
    - You forget the render function: `output$greeting <- "Hello"`
    - You attempt to read from an output: ` message("The greeting is ", output$greeting)`
    
## 4.2 Reactive programming

 - Imperative vs declarative programming
    - Imperative progr: specific command, carried out immediately (loading data, wranfling, plotting, saving...)
    - Decrative progr: higher-level goals or important constraints (shiny)
 - Laziness: app wont run if mispelling (ne error flag)! Double check that your UI and server functions are using the same identifiers.
 - The reactive graph: *gtg* has a reactive dependency on *name*. see reactlog (ch14)
 - Reactive expressions: tool that reduces duplication in your reactive code by introducing additional nodes into the reactive graph `rct_expr()`.
 - Execution order:
    - The order in which reactive code is run is determined only by the reactive graph, not by its layout in the server function.
    - Make sure that reactive expressions and outputs only refer to things defined above, not below11. This will make your code easier to understand.
 
## 4.3 Reactive expressions

  - **producers** will refer to reactive inputs and expressions, 
  - **consumers** will refer to reactive expressions and outputs

### 4.3.1 Motivation
  - compare two simulated datasets with a plot and a hypothesis test. 
  - create 2 costum functions to sue: `freqpoly()` and `t_test()` saved in `help4func1.R`
  - visualise the 2 dist and compare means and summarises the results with a string

### 4.3.2 The app: `app4rctexpr.R`

### 4.3.3 The reactive graph

the graph is very dense: almost every input is connected directly to every output. This creates two problems

  - hard to understand because there are so many connections
  - inefficient because it does more work than necessary
  
these issues can fix all these problems by using reactive expressions to pull out repeated computation

### 4.3.4 Simplifying the graph

  - The “**rule of three**” of programming: whenever you copy and paste something three times, you should figure out how to reduce the duplication (typically by writing a function).
  - In Shiny, however, I think you should consider the rule of one: whenever you copy and paste something once, you should consider extracting the repeated code out into a reactive expression.


```{r}
server <- function(input, output, session) {
  x1 <- reactive(rnorm(input$n1, input$mean1, input$sd1))
  x2 <- reactive(rnorm(input$n2, input$mean2, input$sd2))

  output$hist <- renderPlot({
    freqpoly(x1(), x2(), binwidth = input$binwidth, xlim = input$range)
  }, res = 96)

  output$ttest <- renderText({
    t_test(x1(), x2())
  })
}
```

In the server function below we refactor the existing code to pull out the repeated code into two new reactive expressions, x1 and x2, which simulate the data from the two distributions. To create a reactive expression, we call reactive() and assign the results to a variable. To later use the expression, we call the variable like it’s a function.

### 4.3.5 Why do we need reactive expressions?

When you first start working with reactive code, you might wonder why we need reactive expressions. Why can’t you use your existing tools for reducing duplication in code: creating new variables and writing functions? Unfortunately neither of these techniques work in a reactive environment.


## 4.4 Controlling timing of evaluation: `app4tmg.R`

### 4.4.1 Timed invalidation

`reactiveTimer()` is a reactive expression that has a dependency on a hidden input: the current time. You can use a `reactiveTimer()` when you want a reactive expression to invalidate itself more often than it otherwise would.

### 4.4.2 On click 
 - `actionButton()`: x1() and x2() will update when we click the simulate button, but they’ll also continue to update when lambda1, lambda2, or n change. 
 - `eventReactive()`: 
    - used to replace the existing dependencies, not add to them.
    - x1 and x2 no longer have a reactive dependency on lambda1, lambda2, and n: changing their values will not trigger computation.

## 4.5 Observers: see app4svr.R

  - `observeEvent()`: important debugging tool
  - Use an **observer** to execute actions:
    - saving a file to a shared network drive, 
    - sending data to a web API, 
    - updating a database, or 
    - printing a debugging message to the console (most commonly)
  - There are multiple ways to create an observer: (2 arguments)
    - `eventExpr()`: input or expression to take a dependency on.
    - `handlerExpr()`: code that will be run. 
  - For example, the following modification to `server()` means that every time that *name* is updated, a message will be sent to the console.
  - 2 important differences between `observeEvent()` and `eventReactive()`:
    - You don’t assign the result of `observeEvent()` to a variable, 
    - You can’t refer to it from other reactive consumers.
Observers and outputs are closely related. You can think of outputs as having a special side-effect: updating the HTML in the user’s browser. 



# 5. Case study: ER injuries `app4eri.R`

Exploring data from the National Electronic Injury Surveillance System (NEISS), collected by the Consumer Product Safety Commission. 

This is a long-term study that records all accidents seen in a representative sample of hospitals in the United States. 

More about this dataset [here](https://github.com/hadley/neiss).

Each row represents a single accident with 10 variables:

  - trmt_date: iadmission date.
  - 3 demogr: age, sex, and race 
  - body_part: location of the injury on the body 
  - location: place where the accident occurred (like home or school).
  - diag: basic diagnosis of the injury (like fracture or laceration).
  - prod_code: is the primary product associated with the injury.
  - weight: statistical weight (estimated # people who would suffer this injury in US.
  - narrative: is a brief story about how the accident occurred.
  


Lessons learned:
 - using a custom function for repeated data manipulation `mytop5()`.
 - using `forcats` functions to manipulate categorical data
 - ui layout
 
 
 